JD notes on effects

### 2022-03-23 manuscript/vareffects_plosone.tex
Spell-check

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \cite{hanmer2013behind}. This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}. In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the presponse to be at a particular value or level of the predictor, for an ``average case''. Formerly, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, use the average of the non-focal predictors as the reference point. However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting predictors) versus product of the averages of the interacting predictors (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The grey points are the binned simulated $y$. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. B: Even with the simple interaction between the non-focal predictors, we start seeing deviation of the predicted mean, $\bar{\hat{y}}$, from the simulated, $\bar{y}$, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

@@

For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., model center, in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the focal predictor. For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.

@@

\subsection*{Population-based approach for bias correction}

@@

Population-based approach is not entirely foolproof. For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population. In addition, whole-sample-based approach can be very computationally intensive for large datasets.

>>>>

The most common way of dealing with the first challenge is taking unique levels of the focal (predictor of interest) predictor if discrete or taking appropriately sized quantiles (or bins) if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \cite{hanmer2013behind}. This generates -- \emph{predictor effects}\cite{fox2009effect}, \emph{marginal predictions} \cite{leeper2017package} or \emph{estimated marginal means} \cite{lenth2018package}. In this article, we refer to this quantity as predictor effects since it should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''. More specifically, predictor effects computes the expected outcome by meaningfully holding the non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the outcome expected prediction represents how the model responds to the changes in the focal predictor.

The commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects}) for generating predictor effects, by default, use the average of the non-focal predictors as the reference point. However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging product of interacting model variables) versus product of the averages of the interacting model variables (default for \pkg{emmeans} and \pkg{effects}). We claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ and $\beta_{23} = 5$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

\begin{figure}[!h]
\centering
\includegraphics[width=0.9\textwidth]{justify_preds.inter.pdf}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictor effects for $x_1$ on $y$ for models with and without interactions.}
The horizontal blue, green and black lines are the mean predictions, i.e., $\bar{\hat{y}}$, the red ones are the means of the simulated $y$, i.e., $\bar{y}$, while the vertical dotted grey lines are the means of the corresponding focal predictors. The grey points are the binned simulated $y$. The trend lines represents the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other predictors at their average, based on observed values (red) and the three methods. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the predictions based on the three approaches closely matches the ``true'' predictions. B: Even with the simple interaction between the non-focal predictors, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and ``true'' predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not, the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}

@@

For \pkg{emmeans} and \pkg{effects}, the confidence bands are much wider because they include uncertainties associated with the non-focal predictors, but narrower and crosses at the mean of the focal predictor, i.e., model center, in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate isolated confidence bands indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., mean of the model variable corresponding to the focal predictor. For simple models, the point where the confidence bands crosses is the model center and it corresponds to anchor we choose, in this case, the mean of the focal predictor.

@@

\subsection*{Whole-sample-based approach for bias correction}

@@

Whole-sample-based approach is not entirely foolproof. For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population. In addition, whole-sample-based approach can be very computationally intensive for large datasets.
